Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-14 _tex/paper2.tex
Built site for gh-pages

\begin{equation}\phantomsection\label{eq-mat3}{
RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
}\end{equation}

This matrix also consists of scalars from the set of natural numbers
\(\mathbb{N}\) and has a dimension of
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
columns. The matrix's dimension results from the conceptual-population
model's assumption that researchers have access to data capturing all
repeated judgments made by every available judge for each pair of
stimuli produced by each pair of individuals in the population. Here,
the term \({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial
coefficient, which quantifies the total number of individual-stimuli
comparison pairs within the conceptual population. The coefficient is
defined as follows:

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} \\
\end{aligned}
\]

@@

Figure~\ref{fig-cj11} illustrates the sample-comparison model,
incorporating the sample mechanism. Following the convention of
\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj11_dag} shows
several conditioned variables \((S_{R}\), \(S_{P}\), \(S_{U}\), and
\(S_{V})\) that causally influence the sample-comparison outcome
\(O^{sc}_{RPUV}\), reflecting how the outcome is a sample drawn from the
conceptual population. The DAG also depicts the conceptual-population
outcome \(O^{cp}_{RPUV}\) is now an unobserved variable, emphasizing the
researchers' inability to access such an outcome.

@@

\[
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} \\
\end{aligned}
\]

@@

\subcaption{\label{fig-cj11_scm}SCM}

@@

\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_11.png}

@@

\subcaption{\label{fig-cj11_dag}DAG}

@@

\caption{\label{fig-cj11}Sample-comparison model including sampling
mechanism}

@@

\[
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} \\
\end{aligned}
\]

@@

\subcaption{\label{fig-cj13_scm}SCM}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_13.png}

@@

\subcaption{\label{fig-cj13_dag}DAG}

@@

\caption{\label{fig-cj13}Sample-comparison model including sample and
comparison mechanisms}

@@

>>>>

\begin{equation}\phantomsection\label{eq-mat3}{
RPUV_{n \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
}\end{equation}

This matrix also consists of scalars from the set of natural numbers
\(\mathbb{N}\) and has a dimension of
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and
\(6\) columns. The matrix's dimension results from the
conceptual-population model's assumption that researchers have access to
data capturing all repeated judgments made by every available judge for
each pair of stimuli produced by each pair of individuals in the
population. Here, the term \({n_{I} \cdot n_{A} \choose 2}\) represents
the Binomial coefficient, which quantifies the total number of
individual-stimuli comparison pairs within the conceptual population.
The coefficient is defined as follows:

@@

\[
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} 
\end{aligned}
\]

@@

Figure~\ref{fig-cj12} illustrates the sample-comparison model,
incorporating the sample mechanism. Following the convention of
\citet[pp.~499-516]{McElreath_2020}, the DAG~\ref{fig-cj12_dag} shows
several conditioned variables \((S_{R}\), \(S_{P}\), \(S_{U}\), and
\(S_{V})\) that causally influence the sample-comparison outcome
\(O^{sc}_{RPUV}\), reflecting how the outcome is a sample drawn from the
conceptual population. The DAG also depicts the conceptual-population
outcome \(O^{cp}_{RPUV}\) is now an unobserved variable, emphasizing the
researchers' inability to access such an outcome.

Concatenating the matrices

@@

\[
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj12_scm}SCM}

@@

\includegraphics[width=0.84\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_12.png}

@@

\subcaption{\label{fig-cj12_dag}DAG}

@@

\caption{\label{fig-cj12}Sample-comparison model including sampling
mechanism}

@@

\[
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{RPUV}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{IA}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{JK}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J} 
\end{aligned}
\]

@@

\subcaption{\label{fig-cj14_scm}SCM}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/CJ_TM_14.png}

@@

\subcaption{\label{fig-cj14_dag}DAG}

@@

\caption{\label{fig-cj14}Sample-comparison model including sample and
comparison mechanisms}

@@

\[
\begin{aligned}
  O_{RPUV} & = O^{sc}_{RPUV} \odot C \\
  O^{sc}_{RPUV} & = O^{cp}_{RPUV} \odot S_{RPUV} \\
  O^{cp}_{RPUV} & \sim \text{Bernoulli} \left[ \text{inv\_logit}( D_{RPUV} ) \right] \\
  D_{RPUV} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{IA} & \sim \text{Normal}(0, p_{IA}) \\  
  e_{I} & \sim \text{Normal}(0,1) \\
  e_{JK} & \sim \text{Normal}(0, p_{JK}) \\
  e_{J} & \sim \text{Normal}(0,1)
\end{aligned}
\]

\subcaption{\label{fig-cj15_stat}Statistical model}

\[
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{R}, S_{P}, S_{U}, S_{V}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{IA} & \:\bot\:\{ e_{I}, e_{JK}, e_{J} \} \\
  e_{I} & \:\bot\:\{ e_{JK}, e_{J} \} \\
  e_{JK} & \:\bot\:e_{J}
\end{aligned}
\]

\subcaption{\label{fig-cj15_scm}SCM}

\caption{\label{fig-cj15}Sample-comparison model including sample and
comparison mechanisms}
